---
title: "Data Tidying & Analysis"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(p8105.datasets)
knitr::opts_chunk$set(
	fig.width = 6, 
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Pull in data from Guttmacher and tidy (data from 2014-2017)
```{r}
data = 
  read_csv("./data/112420_guttmacher data.csv")%>%
  select(state_id, measure_name, datum)%>%
  mutate(datum = as.double(datum))%>%
  pivot_wider(names_from = measure_name, values_from = datum)%>%
  janitor:: clean_names()%>%
  rename(public_expenditures = total_reported_public_expenditures_for_family_planning_client_services_in_000s_of_dollars,
         no_women_need_public_expenditure = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_aged_13_44)%>%
  mutate(
    public_expenditures=as.double(public_expenditures),
    no_women_need_public_expenditure=as.double(no_women_need_public_expenditure),
    expenditure_rate = public_expenditures/no_women_need_public_expenditure
  )                                                                                #create expenditure_rate variable
 

```
Load state hostility date (from 2010 data** [ ] complete)
```{r}
hostility =
  read_csv("./data/state_hostility.csv")%>%
  separate('state_id,Hostility',c('state_id', 'hostility'), sep=",")%>%
  mutate(
    hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))
  )
```
### Merge data with hostility rank

### Rename variable names
 [1] `state_id` = "state_id"                                                                                                                                                           
 [2] `counties_no_provider` = "percent_of_counties_without_a_known_abortion_provider"                                                                                                              
 [3] `percent_abortion` = "percent_of_pregnancies_ending_in_abortion"                                                                                                                          
 [4] `percent_birth` = "percent_of_pregnancies_ending_in_birth"                                                                                                                             
 [5] `percent_medicaid` = "percent_of_women_aged_15_44_covered_by_medicaid"                                                                                                                    
 [6] `percent_women_no_provider` = "percent_of_women_aged_15_44_living_in_a_county_without_an_abortion_provider"                                                                                        
 [7] `percent_uninsured` = "percent_of_women_aged_15_44_who_are_uninsured"                                                                                                                      
[8] `percent_bc_18_49` = "percent_of_women_aged_18_49_using_contraceptives"                                                                                                                   
[9] `abortion_rate_15_17_state` = "abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_17_by_state_of_residence"                                                                                 
[10] `abortion_rate_15_19_state` = "abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_19_by_state_of_residence"                                                                                 
[11] `abortion_rate_15_44_state` = "abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_44_by_state_of_residence"                                                                                 
[12] `abortion_rate_18_19_state` = "abortion_rate_the_no_of_abortions_per_1_000_women_aged_18_19_by_state_of_residence"                                                                                 
[13] `birthrate_15_17_state` = "birthrate_the_no_of_births_per_1_000_women_aged_15_17_by_state_of_residence"                                                                                        
[14] `birthrate_15_19_state` = "birthrate_the_no_of_births_per_1_000_women_aged_15_19_by_state_of_residence"                                                                                        
[15] `birthrate_18_19_state` = "birthrate_the_no_of_births_per_1_000_women_aged_18_19_by_state_of_residence"                                                                                        
[16] `need_bc_hisp_20_44` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_hispanic_aged_20_44_and_below_250_percent_of_the_federal_poverty_level"          
[17]  `need_bc_hisp_younger_20` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_hispanic_younger_than_20"                                                        
[18] `need_bc_13_44` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_aged_13_44"                                                                                                                                   
[19] `need_bc_black_20_44` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_black_aged_20_44_and_below_250_percent_of_the_federal_poverty_level"
[20] `need_bc_black_under_20` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_black_younger_than_20"                                              
[21] `need_bc_white_20_44` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_white_aged_20_44_and_below_250_percent_of_the_federal_poverty_level"
[22] `need_bc_white_under_20` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_white_younger_than_20"                                              
[23] `need_bc_under_20` = "no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_younger_than_20"                                                                 
[24] `total_expend_abortion` = "total_reported_public_expenditures_for_abortions_in_000s_of_dollars"                                                                                                
[25] `public_expenditures` = "total_reported_public_expenditures_for_family_planning_client_services_in_000s_of_dollars"                                                                                                                                                
[26] `expenditure_rate` = "public_expenditures"/"need_bc_13_44"                                                                                                                                                 
[27] `hostility` = state hostility ranking in 2000 according to Guttmacher           

```{r}
merge_data = 
  merge(data, hostility, by = "state_id")%>%
  select(-percent_of_pregnancies_wanted_later_or_unwanted_ending_in_abortion, -percent_of_pregnancies_wanted_later_or_unwanted_ending_in_birth
         )%>%
  rename(counties_no_provider = percent_of_counties_without_a_known_abortion_provider,
         percent_abortion = percent_of_pregnancies_ending_in_abortion,
         percent_birth = percent_of_pregnancies_ending_in_birth, 
         percent_medicaid = percent_of_women_aged_15_44_covered_by_medicaid,
         percent_women_no_provider = percent_of_women_aged_15_44_living_in_a_county_without_an_abortion_provider, 
         percent_uninsured = percent_of_women_aged_15_44_who_are_uninsured,
         percent_bc_18_49 = percent_of_women_aged_18_49_using_contraceptives,
         abortion_rate_15_17_state = abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_17_by_state_of_residence,
         abortion_rate_15_19_state = abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_19_by_state_of_residence,
         abortion_rate_15_44_state = abortion_rate_the_no_of_abortions_per_1_000_women_aged_15_44_by_state_of_residence,
         abortion_rate_18_19_state = abortion_rate_the_no_of_abortions_per_1_000_women_aged_18_19_by_state_of_residence,
         birthrate_15_17_state = birthrate_the_no_of_births_per_1_000_women_aged_15_17_by_state_of_residence,
         birthrate_15_19_state = birthrate_the_no_of_births_per_1_000_women_aged_15_19_by_state_of_residence,
         birthrate_18_19_state = birthrate_the_no_of_births_per_1_000_women_aged_18_19_by_state_of_residence,
         need_bc_hisp_20_44 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_hispanic_aged_20_44_and_below_250_percent_of_the_federal_poverty_level,
         need_bc_hisp_younger_20 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_hispanic_younger_than_20,
         need_bc_13_44 = no_women_need_public_expenditure,
         need_bc_black_20_44 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_black_aged_20_44_and_below_250_percent_of_the_federal_poverty_level,
         need_bc_black_under_20 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_black_younger_than_20, 
         need_bc_white_20_44 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_white_aged_20_44_and_below_250_percent_of_the_federal_poverty_level,
         need_bc_white_under_20 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_non_hispanic_white_younger_than_20, 
         need_bc_under_20 = no_of_women_who_likely_need_public_support_for_contraceptive_services_and_supplies_younger_than_20,
         total_expend_abortion = total_reported_public_expenditures_for_abortions_in_000s_of_dollars
         )%>%
    separate('abortion_rate_15_19_state',c(NA, 'abortion_rate_15_19_state'), sep=",")%>%
    separate('abortion_rate_15_44_state',c(NA, 'abortion_rate_15_44_state'), sep=",")%>% 
  separate('birthrate_15_19_state',c(NA, 'birthrate_15_19_state'), sep=",")%>%
    mutate(
    abortion_rate_15_19_state = str_replace_all(abortion_rate_15_19_state, "\\)", ""),
    abortion_rate_15_44_state = str_replace_all(abortion_rate_15_44_state, "\\)", ""),
    birthrate_15_19_state = str_replace_all(birthrate_15_19_state, "\\)", "")
    )%>%
    mutate(
    abortion_rate_15_19_state = as.numeric(abortion_rate_15_19_state),
    abortion_rate_15_44_state = as.numeric(abortion_rate_15_44_state),
    birthrate_15_19_state = as.numeric(birthrate_15_19_state),
    counties_no_provider = as.numeric(counties_no_provider),
    percent_abortion = as.numeric(percent_abortion),
   percent_birth = as.numeric(percent_birth),
    percent_medicaid = as.numeric(percent_medicaid),
    percent_women_no_provider = as.numeric(percent_women_no_provider),
    percent_uninsured = as.numeric(percent_uninsured),
    percent_bc_18_49 = as.numeric(percent_bc_18_49),
    abortion_rate_15_17_state = as.numeric(abortion_rate_15_17_state),
    abortion_rate_15_19_state = as.numeric(abortion_rate_15_19_state),
    abortion_rate_18_19_state = as.numeric(abortion_rate_18_19_state),
    birthrate_15_17_state = as.numeric(birthrate_15_17_state),
     birthrate_18_19_state = as.numeric( birthrate_18_19_state),
    need_bc_hisp_20_44 = as.numeric(need_bc_hisp_20_44),
    need_bc_hisp_younger_20 = as.numeric(need_bc_hisp_younger_20),
     need_bc_black_20_44 = as.numeric( need_bc_black_20_44),
    need_bc_black_under_20 = as.numeric(need_bc_black_under_20),
   need_bc_white_20_44 = as.numeric(need_bc_white_20_44),
    need_bc_white_under_20 = as.numeric(need_bc_white_under_20),
    need_bc_under_20  = as.numeric(need_bc_under_20),
   total_expend_abortion  = as.numeric(total_expend_abortion)
    )
```

```{r}
write_csv(merge_data, "./data/merge_data.csv")
```

