---
title: "Factors Influencing Abortion Rates in the US"
author: "Linh, Maggie, Nancy, Rachel"
date: "December 5, 2020"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r include = FALSE}
library(tidyverse)
library(leaflet)
library(sf)
library(rgdal)
library(huxtable)
library(patchwork)
library(broom)
library(plotly)
library(devtools)
library(modelr)
library(mgcv)

knitr::opts_chunk$set(
	fig.width = 6, 
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Motivation
For this project, we aim to examine state-level sociodemographic factors that may influence abortion rate. State family planning policies vary widely, with some states more hostile to women attempting to access an abortion, and others more supportive (Nash 2019). One study showed that of women who received a nonhospital abortion, 24% traveled at least 50 miles from their home to an abortion facility (Henshaw 1995). We plan to investigate the relationship between hostility towards abortion, within-county abortion provider access, and abortion rate.  We are interested in visualizing these data to hopefully help disseminate this information in a digestible way to inform outreach/education strategies and policy.

# Related Work
We are additionally interested in the relationship between teen birth rate, public expenditures for contraception, and hostility. 

# Research Questions
* Is there a relationship between hostility towards abortion and access to abortion?
* Is there a relationship between hostility towards abortion and abortion rates by state?
* Is there a relationship between abortion access and abortion rates by state?
* Is there a relationship between abortion access and hostility towards abortion with abortion rates by state?

# Data Sources

The data are pulled from the Guttmacher Institute, which is a research organization founded in 1968 working to study, educate, and advance sexual and reproductive health and rights. The Guttmacher Institute Data Center provides public-access birth, abortion, and contraceptive use demographic data at the national and state levels, and contraceptive use data only at the county level. Since we are interested in multiple study outcomes—abortion rates, contraceptive funding/access, and teen birth rates—we pulled these data at the finest resolution possible, aggregated at the state level, for all 50 US States.

# Exploratory Analysis

```{r include = FALSE}
merge_data = read_csv("data/merge_data.csv")
```

## Scatterplot comparing abortion rates and public expenditure on contraceptives 

```{r, fig.height = 12, fig.width = 6}
merge_data %>% 
 mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% 
 mutate(text_label = str_c("State: ", state_id, "\nHostility: ", hostility)) %>% 
  plot_ly(x = ~abortion_rate_15_44_state, y = ~expenditure_rate, color = ~hostility, type = "scatter", colors = "viridis", size = 3, text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Number of abortions per 1000 women aged 15-44"),
    yaxis = list(title = "Expenditure Rate")
  )

```


## Rate of abortion by state

```{r, fig.height = 6, fig.width = 12, fig.asp = .6}
merge_data %>% 
  mutate(state_id = fct_reorder(state_id, abortion_rate_15_44_state)) %>% 
  ggplot(aes(x = state_id, y = abortion_rate_15_44_state, fill = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Abortion rate",
    title = "Rate of abortion per 1000 women aged 15-44 by state of residence"
  ) +
  theme(legend.position = "none")
```

## Public expenditure by state

```{r, fig.height = 6, fig.width = 12, fig.asp = .6}
merge_data %>% 
  mutate(state_id = fct_reorder(state_id, expenditure_rate)) %>% 
  ggplot(aes(x = state_id, y = expenditure_rate, fill = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Public expenditure per women in need",
    title = "Rate of public expenditure for women aged 15-44 in need of contraception"
  ) +
  theme(legend.position = "none")
  
```

## Abortion Rate and Public Expenditure -- 

```{r, fig.height = 12, fig.width = 12, fig.asp = .6}
abortion_rate = 
  merge_data %>% 
  ggplot(aes(x = state_id, y = abortion_rate_15_44_state, color = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Abortion rate",
    title = "Rate of abortion per 1000 women aged 15-44 by state of residence"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) 
 

public_expenditure =
  merge_data %>% 
  ggplot(aes(x = state_id, y = expenditure_rate, color = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Public expenditure per women in need",
    title = "Rate of public expenditure for women aged 15-44 in need of contraception"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

abortion_rate / public_expenditure
```

## Hostility towards abortion in policy by state and abortion rate

```{r, fig.height = 9, fig.width = 9, fig.asp = .6}
merge_data %>% 
  select(state_id, percent_abortion, abortion_rate_15_44_state ,hostility) %>% 
  mutate(hostility = as.factor(hostility)) %>% 
  group_by(hostility) %>% 
  summarize(
    n = n(),
    abortion_rate = mean(abortion_rate_15_44_state)
  ) %>% 
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% 
  ggplot(aes(x = hostility, y = abortion_rate, fill = hostility)) +
  geom_bar(stat = "identity") + 
  labs(
    x = "Hostility towards abortion in policy by state",
    y = "Average rate of abortion by state",
    title = "Hostility and Abortion Rate"
  )
  
```

## Spatial Visualizations

To create the maps of our relevant variables, we first read in a US state shapefile downloaded from the [US Census Bureau](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html). We then read in our tidied Guttmacher dataset, and did some additional tidying.

```{r}
# import US state shapefile from data folder
all_states <- readOGR("data/cb_2018_us_state_500k/cb_2018_us_state_500k.shp")

# read in project data
merge_data = read_csv("data/merge_data.csv") %>% 
  rename(STUSPS = state_id) %>% # to match shapefile column name 
  mutate_if(is.numeric, round, digits = 2) %>%  # round to 2 decimal places
  mutate(hostility = as.factor(hostility)) %>%
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% # for mapping hostility as a categorical variable
  mutate(hostility = recode(hostility,
                            hostile = "Hostile",
                            leans_hostile = "Leans Hostile",
                            middle_ground = "Middle Ground",
                            leans_supportive = "Leans Supportive",
                            supportie = "Supportive")) # cleaner hostility names for map

```

We then joined the Guttmacher data to the shapefile.

```{r}
data_sf <- merge(all_states, merge_data, all.x = F) 
```

We then created a universal basemap and for each map, we created a continuous-scale color palette.

**Abortion Rates**

```{r}
# make color palette
abortion_rate_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$percent_abortion,
  reverse = TRUE)

# base map
base_map <- leaflet(data_sf) %>%
  addProviderTiles("CartoDB.PositronNoLabels") 

# add in custom labels for percent abortion
labels_abortion_rate <- sprintf(
  "<strong>%s</strong><br/>%g&#37;",
  data_sf$STUSPS, data_sf$percent_abortion
) %>% lapply(htmltools::HTML)

# full map
abortion_map = base_map %>% 
  addPolygons(
  fillColor = ~abortion_rate_pal(data_sf$percent_abortion),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_abortion_rate,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = abortion_rate_pal, values = ~percent_abortion, opacity = 0.7, title = 'Abortion Rates by State <br> (% of pregnancies ending in abortions)',
  position = "bottomright")
abortion_map
```

**Public Contraceptive Expenditure Rates**

```{r}
# make color palette
bc_fund_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$expenditure_rate,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_bc_fund <- sprintf(
  "<strong>%s</strong><br/>&#36;%g/woman",
  data_sf$STUSPS, data_sf$expenditure_rate
) %>% lapply(htmltools::HTML)

# full map
bc_fund_map = base_map %>% 
  addPolygons(
  fillColor = ~bc_fund_pal(data_sf$expenditure_rate),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_bc_fund,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = bc_fund_pal, values = ~expenditure_rate, opacity = 0.7, title = 'Public Contraceptive Expenditure Rate by State  <br> ($/woman in likely need of publicly-funded services)',
  position = "bottomright")

bc_fund_map
```

**Percent of Women Without Access to Abortion Provider in County of Residence**

```{r}
# make color palette
abortion_access_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$percent_women_no_provider,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_abortion_access <- sprintf(
  "<strong>%s</strong><br/>%g&#37;",
  data_sf$STUSPS, data_sf$percent_women_no_provider
) %>% lapply(htmltools::HTML)

# full map
abortion_access_map = base_map %>% 
  addPolygons(
  fillColor = ~abortion_access_pal(data_sf$percent_women_no_provider),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_abortion_access,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = abortion_access_pal, values = ~percent_women_no_provider, opacity = 0.7, title = 'Percent of women living in a county <br> without access to abortion clinic by State',
  position = "bottomright")

abortion_access_map
```

**Teen Birth Rates**

```{r}
# make color palette
teen_births_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$birthrate_15_19_state,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_teen_births <- sprintf(
  "<strong>%s</strong><br/> %g births per 1000 women",
  data_sf$STUSPS, data_sf$birthrate_15_19_state
) %>% lapply(htmltools::HTML)


# full map
teen_births_map = base_map %>% 
  addPolygons(
  fillColor = ~teen_births_pal(data_sf$birthrate_15_19_state),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_teen_births,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = teen_births_pal, values = ~birthrate_15_19_state, opacity = 0.7, title = 'No. of births per 1000 women aged 15-19',
  position = "bottomright")

teen_births_map
```

**Hostility toward Abortion Rights by State in 2010**

```{r}
# make color palette
hostility_pal <- colorFactor(
  palette = "inferno",
  domain = data_sf$hostility,
  reverse = F)

# add in custom labels for public contraceptive expenditure
labels_hostility <- sprintf(
  "<strong>%s</strong><br/>%s",
  data_sf$STUSPS, data_sf$hostility
) %>% lapply(htmltools::HTML)

# full map
hostility_map = base_map %>% 
  addPolygons(
  fillColor = ~hostility_pal(data_sf$hostility),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_hostility,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = hostility_pal, 
            values = ~hostility, 
            opacity = 1, 
            title = 'Hostility Rating towards Abortion Rights in 2010',
  position = "bottomright")

hostility_map
```

## RACHEL Map ShinyApp

# MAGGIE Regression Models

We ran a few different linear regression analyses to round out our exploratory data analysis. The tables below summarizes the effect estimates we found for these models:

1. Simple linear regression: Is there a relationship between hostility towards abortion and access to abortion?

```{r}
hostility_bc_access <- lm(percent_women_no_provider ~ hostility, data = merge_data) 
hostility_bc_access_tbl = hux(
  Outcome = c("Outcome", 
              "Abortion Access", "", "", "", ""), 
  Predictors = c("Hostility Rating", 
                 "Hostile (Intercept)", 
                 "Leans Hostile", 
                 "Middle Ground", 
                 "Leans Supportive", 
                 "Supportive"), 
  Estimate = c("Estimate", 
               hostility_bc_access$coefficients[1], 
               hostility_bc_access$coefficients[2],
               hostility_bc_access$coefficients[3],
               hostility_bc_access$coefficients[4],
               hostility_bc_access$coefficients[5]), 
  P_value = c("P-value", 
              summary(hostility_bc_access)$coefficients[,"Pr(>|t|)"][1], 
              summary(hostility_bc_access)$coefficients[,"Pr(>|t|)"][2],
              summary(hostility_bc_access)$coefficients[,"Pr(>|t|)"][3],
              summary(hostility_bc_access)$coefficients[,"Pr(>|t|)"][4],
              summary(hostility_bc_access)$coefficients[,"Pr(>|t|)"][5]),
  add_colnames = FALSE
)
hostility_bc_access_tbl %>% 
  set_bold(1, everywhere, TRUE) %>% 
  set_bottom_border(1, everywhere, 2) %>% 
  set_align(everywhere, 2, 'right') %>% 
  set_right_padding(10) %>% 
  set_left_padding(10) %>% 
  set_width(0.35) %>% 
  set_number_format(3)

```


2. Simple linear regression: Is there a relationship between hostility towards abortion and abortion rates by state?

```{r}
hostility_model = lm(abortion_rate_15_44_state ~ hostility, data = merge_data) 

```

3. Simple linear regression: Is there a relationship between abortion access and abortion rates by state?

```{r}
bc_access_model <- lm(abortion_rate_15_44_state ~ percent_women_no_provider, data = merge_data) 



```

# Multivariate regression: Is there a relationship between abortion access and hostility towards abortion with abortion rates by state?

```{r}
hostility_adj_model = lm(abortion_rate_15_44_state ~ hostility + percent_women_no_provider, data = merge_data)
```


# ALL Discussion and Conclusion