---
title: "Maps"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source_code: embed
runtime: shiny
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(flexdashboard)
library(shiny)
library(plotly)
library(tidyverse)

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# read in project data
merge_data = read_csv("./data/merge_data.csv") %>% 
  mutate_if(is.numeric, round, digits = 2) %>%  # round to 2 decimal places
  select(state_id, percent_abortion, percent_women_no_provider, expenditure_rate, birthrate_15_19_state) %>% 
  rename(
    state = state_id,
    "% pregnancies ending in abortion" = percent_abortion,
    "% living in a county with no abortion provider" = percent_women_no_provider,
    "Contraceptive Expenditure Rate" = expenditure_rate,
    "Teen Birth Rate" = birthrate_15_19_state
  ) %>% 
  pivot_longer("% pregnancies ending in abortion":"Teen Birth Rate",
    names_to = "type_variable",
    values_to = "Measure"
  )

```


Column {.sidebar}
-----------------------------------------------------------------------

```{r echo = FALSE}
type_variable <- (merge_data %>% distinct(type_variable) %>% pull())

selectInput("variable_type", label = h3("Family planning variable"),
            choices = type_variable, selected = "Abortion Rate")

```


Column {.tabset}
-----------------------------------------------------------------------

### US MAP

```{r echo=FALSE}
renderPlotly({
  geo1 <- list(
    scope = "usa",
    projection = list(type =  "state"),
    showlakes = TRUE,
    lakecolor = toRGB("white")
  )
  
  merge_data %>% 
    filter(type_variable == input$variable_type) %>% 
    plot_geo(locationmode = "USA-states") %>% 
    add_trace(
      z = ~Measure,
      locations = ~state,
      color = ~Measure,
      colors = "inferno",
      reversescale = TRUE,
      text = ~paste(type_variable)
    ) %>% 
    layout(
      geo = geo1,
      title = "Family Planning Statistics by State",
      legend = list(x = 100, y = 0.5)
    )
})


```

### Variable Descriptions

**% pregnancies ending in abortion**
Percent of pregnancies in a given state that ended in abortion

**% living in a county with no abortion provider**
Percent of women aged 15-44 living in a county that does not have an abortion provider

**Contraceptive Expenditure Rate**
Total state public funding for contraception divided by the number of women aged 13-44 in likely need of contraceptive services and supplies

**Teen Birth Rate**
Number of births per 1000 women aged 15-19

```{r eval=FALSE, include=FALSE}
#rsconnect::deployApp("./maps.Rmd")
```
