---
title: "Regression analysis"
author: "Linh Tran"
date: "11/30/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(patchwork)
library(broom)
library(plotly)
library(devtools)
library(leaflet)
library(modelr)
library(mgcv)

knitr::opts_chunk$set(
	fig.width = 6, 
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Load dataset

```{r}
merge_data = read_csv("data/merge_data.csv")
```

# Exploratory data analysis

## Rate of abortion and public contraceptive expenditure by state?

```{r, fig.height = 12, fig.width = 6}
merge_data %>% 
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% 
  ggplot(aes(x = abortion_rate_15_44_state, y = expenditure_rate, color = hostility)) +
  geom_point(size = 3)
```


## Rate of abortion by state

```{r, fig.height = 6, fig.width = 12, fig.asp = .6}
merge_data %>% 
  mutate(state_id = fct_reorder(state_id, abortion_rate_15_44_state)) %>% 
  ggplot(aes(x = state_id, y = abortion_rate_15_44_state, fill = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Abortion rate",
    title = "Rate of abortion per 1000 women aged 15-44 by state of residence"
  ) +
  theme(legend.position = "none")
```

## Public expenditure by state

```{r, fig.height = 6, fig.width = 12, fig.asp = .6}
merge_data %>% 
  mutate(state_id = fct_reorder(state_id, expenditure_rate)) %>% 
  ggplot(aes(x = state_id, y = expenditure_rate, fill = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Public expenditure per women in need",
    title = "Rate of public expenditure for women aged 15-44 in need of contraception"
  ) +
  theme(legend.position = "none")
  
```


```{r, fig.height = 12, fig.width = 12, fig.asp = .6}
abortion_rate = 
  merge_data %>% 
  ggplot(aes(x = state_id, y = abortion_rate_15_44_state, color = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Abortion rate",
    title = "Rate of abortion per 1000 women aged 15-44 by state of residence"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) 
 

public_expenditure =
  merge_data %>% 
  ggplot(aes(x = state_id, y = expenditure_rate, color = state_id)) +
  geom_bar(stat = "identity") +
  labs(
    x = "State",
    y = "Public expenditure per women in need",
    title = "Rate of public expenditure for women aged 15-44 in need of contraception"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

abortion_rate / public_expenditure
```

## Hostility towards abortion in policy by state and abortion rate

```{r, fig.height = 9, fig.width = 9, fig.asp = .6}
merge_data %>% 
  select(state_id, percent_abortion, abortion_rate_15_44_state ,hostility) %>% 
  mutate(hostility = as.factor(hostility)) %>% 
  group_by(hostility) %>% 
  summarize(
    n = n(),
    abortion_rate = mean(abortion_rate_15_44_state)
  ) %>% 
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% 
  ggplot(aes(x = hostility, y = abortion_rate, fill = hostility)) +
  geom_bar(stat = "identity") + 
  labs(
    x = "Hostility towards abortion in policy by state",
    y = "Average rate of abortion by state",
    title = "Hostility and Abortion Rate"
  )
  
```


## Top 10 states with the highest abortion rate

```{r}
top_abortion_rate =
  merge_data %>% 
  top_n(10, abortion_rate_15_44_state) %>% 
  mutate(state_id = fct_reorder(state_id, abortion_rate_15_44_state)) %>% 
  plot_ly(x = ~abortion_rate_15_44_state, y = ~state_id, type = 'bar', orientation = 'h', width = 900, height = 700, marker = list(color = c('rgba(222,45,38,1)', 'rgba(222,45,38,1)', 'rgba(222,45,38,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(222,45,38,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)'))) %>% 
  layout(title = "Top 10 states with highest abortion rate",
         xaxis = list(title = "Abortion rate"),
         yaxis = list(title = "State"),
         showlegend = FALSE)

top_expend_abortion = 
  merge_data %>% 
  top_n(10, total_expend_abortion) %>% 
  mutate(state_id = fct_reorder(state_id, total_expend_abortion)) %>% 
  plot_ly(x = ~total_expend_abortion, y = ~state_id, type = 'bar', orientation = 'h', width = 900, height = 700, marker = list(color = c('rgba(222,45,38,1)', 'rgba(222,45,38,1)', 'rgba(204,204,204,1)',  'rgba(222,45,38,1)',  'rgba(222,45,38,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)'))) %>% 
  layout(title = "Top 10 states with highest total expend",
         xaxis = list(title = "total_expend_abortion"),
         yaxis = list(title = "State"),
         showlegend = FALSE)

subplot(top_abortion_rate, top_expend_abortion, shareX = FALSE, shareY = FALSE,
        titleX = FALSE, titleY = FALSE)
  
```


## Statistical analysis

### Is there a relationship between rate of abortion and hostility towards abortion in policy by state and/or contraceptive funding/abortion access by state? **ALL MODELS HERE**

```{r}
merge_data = 
  merge_data %>% 
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive")))

# simple linear regression: is there a relationship between hostility towards abortion and access to abortion?
hostility_bc_access <- lm(percent_women_no_provider ~ hostility, data = merge_data)
summary(hostility_bc_access)

# simple linear regression: is there a relationship between hostility towards abortion and abortion rates by state?
hostility_model = lm(abortion_rate_15_44_state ~ hostility, data = merge_data)
summary(hostility_model)

# simple linear regression: is there a relationship between abortion access and abortion rates by state?
bc_access_model <- lm(abortion_rate_15_44_state ~ percent_women_no_provider, data = merge_data) 
summary(bc_access_model)

# multivariate regression: is there a relationship between abortion access and hostility towards abortion with abortion rates by state?
hostility_adj_model = lm(abortion_rate_15_44_state ~ hostility + percent_women_no_provider, data = merge_data)

summary(hostility_adj_model)

```

### Is there a relationship between rate of abortion and public contraceptive expenditure by state? **probably not going to use**

```{r}
bc_fund_model <- lm(abortion_rate_15_44_state ~ expenditure_rate, data = merge_data)
summary(bc_fund_model)
```

### Is there a relationship between rate of abortion and abortion access by state? **moved to chunk above**

```{r}
bc_access_model <- lm(abortion_rate_15_44_state ~ percent_women_no_provider, data = merge_data) 
summary(bc_access_model)
```

### Is there a relationship between teen birth rate and percent of women in need of publicly-funded contraceptive resources? **probably not going to use**

```{r}
teenbirth_model <- lm(birthrate_15_19_state ~ need_bc_under_20, data = merge_data) 
summary(teenbirth_model)
```

