---
title: "Analysis"
author: "Linh, Maggie, Nancy, Rachel"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(patchwork)
library(broom)
library(plotly)
library(devtools)
library(leaflet)
library(modelr)
library(mgcv)

knitr::opts_chunk$set(
	fig.width = 6, 
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r include = FALSE}
merge_data = read_csv("data/merge_data.csv")
```

## Statistical analysis

### Is there a relationship between rate of abortion and hostility towards abortion in policy by state and/or contraceptive funding/abortion access by state? **ALL MODELS HERE**

```{r}
merge_data = 
  merge_data %>% 
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive")))

# simple linear regression: is there a relationship between hostility towards abortion and access to abortion?
hostility_bc_access <- lm(percent_women_no_provider ~ hostility, data = merge_data)
summary(hostility_bc_access)

# simple linear regression: is there a relationship between hostility towards abortion and abortion rates by state?
hostility_model = lm(abortion_rate_15_44_state ~ hostility, data = merge_data)
summary(hostility_model)

# simple linear regression: is there a relationship between abortion access and abortion rates by state?
bc_access_model <- lm(abortion_rate_15_44_state ~ percent_women_no_provider, data = merge_data) 
summary(bc_access_model)

# multivariate regression: is there a relationship between abortion access and hostility towards abortion with abortion rates by state?
hostility_adj_model = lm(abortion_rate_15_44_state ~ hostility + percent_women_no_provider, data = merge_data)
summary(hostility_adj_model)

```

### Is there a relationship between rate of abortion and public contraceptive expenditure by state? **probably not going to use**

```{r}
bc_fund_model <- lm(abortion_rate_15_44_state ~ expenditure_rate, data = merge_data)
summary(bc_fund_model)
```

### Is there a relationship between rate of abortion and abortion access by state? **moved to chunk above**

```{r}
bc_access_model <- lm(abortion_rate_15_44_state ~ percent_women_no_provider, data = merge_data) 
summary(bc_access_model)
```

### Is there a relationship between teen birth rate and percent of women in need of publicly-funded contraceptive resources? **probably not going to use**

```{r}
teenbirth_model <- lm(birthrate_15_19_state ~ need_bc_under_20, data = merge_data) 
summary(teenbirth_model)
```

