---
title: "Spatial Visualization of Data"
output: html_document
---

## Simple Univariate Chloropleth Maps using Leaflet

* Hostility rate (although the map already exists for this... so is there a need? maybe we just link it): https://www.guttmacher.org/article/2019/08/state-abortion-policy-landscape-hostile-supportive

#### Load libraries and dataset

```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(rgdal)

# import US state shapefile from data folder
all_states <- readOGR("data/cb_2018_us_state_500k/cb_2018_us_state_500k.shp")
summary(all_states)


# read in project data
merge_data = read_csv("data/merge_data.csv") %>% 
  rename(STUSPS = state_id) %>% # to match shapefile column name 
  mutate_if(is.numeric, round, digits = 2) %>%  # round to 2 decimal places
  mutate(hostility = as.factor(hostility)) %>%
  mutate(hostility = fct_relevel(hostility, c("hostile", "leans_hostile", "middle_ground", "leans_supportive", "supportive"))) %>% # for mapping hostility as a categorical variable
  mutate(hostility = recode(hostility,
                            hostile = "Hostile",
                            leans_hostile = "Leans Hostile",
                            middle_ground = "Middle Ground",
                            leans_supportive = "Leans Supportive",
                            supportie = "Supportive")) # cleaner hostility names for map
merge_data
```

#### Join project data with sf file
```{r}
data_sf <- merge(all_states, merge_data, all.x = F) 
summary(data_sf)
```

#### Abortion rates map

```{r}
# make color palette
abortion_rate_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$percent_abortion,
  reverse = TRUE)

# base map
base_map <- leaflet(data_sf) %>%
  addProviderTiles("CartoDB.PositronNoLabels") 

# add in custom labels for percent abortion
labels_abortion_rate <- sprintf(
  "<strong>%s</strong><br/>%g&#37;",
  data_sf$STUSPS, data_sf$percent_abortion
) %>% lapply(htmltools::HTML)

# full map
abortion_map = base_map %>% 
  addPolygons(
  fillColor = ~abortion_rate_pal(data_sf$percent_abortion),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_abortion_rate,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = abortion_rate_pal, values = ~percent_abortion, opacity = 0.7, title = 'Abortion Rates by State <br> (% of pregnancies ending in abortions)',
  position = "bottomright")
abortion_map
```


#### Public contraceptive expenditure rate map

```{r}
# make color palette
bc_fund_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$expenditure_rate,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_bc_fund <- sprintf(
  "<strong>%s</strong><br/>&#36;%g/woman",
  data_sf$STUSPS, data_sf$expenditure_rate
) %>% lapply(htmltools::HTML)

# full map
bc_fund_map = base_map %>% 
  addPolygons(
  fillColor = ~bc_fund_pal(data_sf$expenditure_rate),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_bc_fund,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = bc_fund_pal, values = ~expenditure_rate, opacity = 0.7, title = 'Public Contraceptive Expenditure Rate by State  <br> ($/woman in likely need of publicly-funded services)',
  position = "bottomright")

bc_fund_map

```

#### Abortion access map

```{r}
# make color palette
abortion_access_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$percent_women_no_provider,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_abortion_access <- sprintf(
  "<strong>%s</strong><br/>%g&#37;",
  data_sf$STUSPS, data_sf$percent_women_no_provider
) %>% lapply(htmltools::HTML)

# full map
abortion_access_map = base_map %>% 
  addPolygons(
  fillColor = ~abortion_access_pal(data_sf$percent_women_no_provider),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_abortion_access,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = abortion_access_pal, values = ~percent_women_no_provider, opacity = 0.7, title = 'Percent of women living in a county <br> without access to abortion clinic by State',
  position = "bottomright")

abortion_access_map
```

#### Teen birth rates map

```{r}
# make color palette
teen_births_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$birthrate_15_19_state,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_teen_births <- sprintf(
  "<strong>%s</strong><br/> %g births per 1000 women",
  data_sf$STUSPS, data_sf$birthrate_15_19_state
) %>% lapply(htmltools::HTML)


# full map
teen_births_map = base_map %>% 
  addPolygons(
  fillColor = ~teen_births_pal(data_sf$birthrate_15_19_state),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_teen_births,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = teen_births_pal, values = ~birthrate_15_19_state, opacity = 0.7, title = 'No. of births per 1000 women aged 15-19',
  position = "bottomright")

teen_births_map
```

#### Number of Women under 20 in need of accessing publicly-funded contraceptives **don't need to include in final report**

```{r}
# make color palette
need_bc_teen_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$need_bc_under_20,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_need_bc_teen <- sprintf(
  "<strong>%s</strong><br/>%g",
  data_sf$STUSPS, data_sf$need_bc_under_20
) %>% lapply(htmltools::HTML)

# full map
need_bc_teen_map = base_map %>% 
  addPolygons(
  fillColor = ~need_bc_teen_pal(data_sf$need_bc_under_20),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_need_bc_teen,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = need_bc_teen_pal, values = ~need_bc_under_20, opacity = 0.7, title = 'No. of women who likely need public support for contraceptive services and supplies younger than 20',
  position = "bottomright")

need_bc_teen_map
```

#### Total Number of Women in need of accessing publicly-funded contraceptives **dont need this**

```{r}
# make color palette
need_bc_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$need_bc_13_44,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_need_bc <- sprintf(
  "<strong>%s</strong><br/>%g",
  data_sf$STUSPS, data_sf$need_bc_13_44
) %>% lapply(htmltools::HTML)

# full map
need_bc_map = base_map %>% 
  addPolygons(
  fillColor = ~need_bc_pal(data_sf$need_bc_13_44),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_need_bc,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = need_bc_pal, values = ~need_bc_13_44, opacity = 0.7, title = 'Total No. of women who likely need public support for contraceptive services and supplies',
  position = "bottomright")

need_bc_map
```

#### State expenditure rate **repeat of map above**

```{r}
# make color palette
expenditure_rate_pal <- colorNumeric(
  palette = "inferno",
  domain = data_sf$expenditure_rate,
  reverse = TRUE)

# add in custom labels for public contraceptive expenditure
labels_expenditure_rate <- sprintf(
  "<strong>%s</strong><br/>%g",
  data_sf$STUSPS, data_sf$expenditure_rate
) %>% lapply(htmltools::HTML)

# full map
expenditure_rate_map = base_map %>% 
  addPolygons(
  fillColor = ~expenditure_rate_pal(data_sf$expenditure_rate),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_expenditure_rate,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = expenditure_rate_pal, values = ~expenditure_rate, opacity = 0.7, title = 'Ratio of Total reported public expenditures for family planning client services in 1000s of dollars to number of women in likely need of public support for contraceptive services and supplies',
  position = "bottomright")

expenditure_rate_map
```

#### Hostility map

```{r eval=FALSE, include=FALSE}
# make color palette
hostility_pal <- colorFactor(
  palette = "inferno",
  domain = data_sf$hostility,
  reverse = F)

# add in custom labels for public contraceptive expenditure
labels_hostility <- sprintf(
  "<strong>%s</strong><br/>%s",
  data_sf$STUSPS, data_sf$hostility
) %>% lapply(htmltools::HTML)

# full map
hostility_map = base_map %>% 
  addPolygons(
  fillColor = ~hostility_pal(data_sf$hostility),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels_hostility,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>% 
  addLegend(pal = hostility_pal, 
            values = ~hostility, 
            opacity = 1, 
            title = 'Hostility Rating towards Abortion Rights in 2010',
  position = "bottomright")

hostility_map
```

